Seabird assemblages are linked to the major western boundary current off eastern Australia

### 2023-10-24 ms_preprint/ms_preprint.Rmd
ms adjustments after LB RS and MS reviews pre-submission

---
title: Seabird assemblages are linked to the major western boundary current off eastern Australia
authors:
  - name: Nicholas W. Daudt*
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: nicholaswdaudt@gmail.com
    notes: Corresponding author
  - name: Eric J. Woehler
    department: Australasian Seabird Group
    affiliation: BirdLife Australia
    location: Hobart, Australia
    email: eric.woehler@gmail.com
  - name: Matthew R. Schofield
    department: Department of Mathematics and Statistics
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: matthew.schofield@otago.ac.nz
  - name: Robert O. Smith
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: robert.smith@otago.ac.nz
  - name: Leandro Bugoni
    department: Institute of Biological Sciences
    affiliation: Universidade Federal do Rio Grande--FURG
    location: Rio Grande, RS, Brazil
    email: lbugoni@yahoo.com.br
  - name: William J. Rayment
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: will.rayment@otago.ac.nz
abstract: |
  **Aim:** To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Location:** East coast of Australia.
  **Methods:** We used 15 seabird at-sea ship-based survey data sets collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.
  **Results:** Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at ~34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Main conclusions:** RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
keywords:
  - Assemblages
  - Biogeography
  - Community ecology
  - East Australian Current
  - Mixture models
  - Region of Common Profiles
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \newcommand{\blinenumbers}{\begin{linenumbers}}
- \newcommand{\elinenumbers}{\end{linenumbers}}
- \usepackage{pdflscape}
- \DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}} # needed to understand unicode for greek Delta in Table 1
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---

<!-- Running title: Seabird assemblages off Eastern Australia -->

<!-- Note: I've changed the 'arxiv.sty' source file to include the running title in the PDF -->

@@

William J. Rayment -- <https://orcid.org/0000-0001-9558-5109>

@@

We thank all observers who assisted in data collection, the crews of RV *Investigator* and RV *Falkor*, and the Australasian Seabird Group (BirdLife Australia) for logistical support. The authors would like to thank Dr. Skipton N.C. Woolley, `{ecomix}` developer, who kindly helped NWD build preliminary models and gave advice in several stages of model construction. NWD is supported by a Doctoral scholarship from the University of Otago.

@@

**Data availability statement:** The gridded data used as the basis for all analyses are available in the archived repository along with the code to reproduce the results. The raw data are currently being used for other projects and thus can not be publicly available. If you are interested in the raw data, please get in touch with Eric J. Woehler.

@@

**Results:** Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages ('northern' and 'southern'), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at \~34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.

**Main conclusions:** RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.

@@

Seabird assemblages have been widely described and related to environmental characteristics [@ribic1997; @woehler2003; @hyrenbach2007]. Considering the hierarchical nature of oceanographic processes and their resulting features [@haury1978patterns], it is not surprising that the distribution of highly mobile seabirds are linked differently across temporal and spatial scales [@haury1978patterns; @hunt1987]. In the marine environment, large scale (\>100 km length) oceanographic features (e.g. boundary currents and water masses) act as terrestrial biomes (based on vegetation), and influence seabird assemblage structure--i.e. the occurrence of species and their relative numerical composition [@ribic1997; @hyrenbach2007; @commins2014]. In contrast, local abundance is related to meso- (10--100 km) and sub-mesoscale (1--10 km) processes that aggregate prey and therefore predators, such as fronts and eddies [@weimerskirch2007; @scales2014]. Such processes, although relatively ephemeral, are primarily driven by physical and atmospheric forces, and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur [@davoren2013; @hazen2013]. Sea surface temperatures are a key driver of seabird distributions, at individual- and assemblage-levels, operating over a range of spatial and temporal scales from macro (e.g. water masses) to fine (e.g. fronts) [@dunlop1988; @ribic1997; @quillfeldt2015]. Despite extensive surveys [e.g. @ballance1997] and tracking studies [e.g. @carneiro2020], many oceanic regions still lack information on the distributions and abundances of seabird species, and how they assemble in space and time.

@@

In this study, we aim to identify assemblages of seabird species within the EAC and adjacent areas, and relate these assemblages to environmental conditions, by using at-sea survey data and a state-of-the-art modelling technique. We use the Region of Common Profiles [RCP; @foster2013], a mixture model regarded as one of the best available tools for describing bioregions [@hill2020; @woolley2020]. RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics [@warton2015; @hill2020]. Eastern Australia supports a high diversity of seabirds and offers wide environmental gradients over a near 40° latitudinal range, from the tropics to the cool temperate. We hypothesised to find different seabird assemblages relating to different environmental conditions within the EAC. In addition, presence-absence and abundance data were used to test if the different numerical resolutions would result in the identification of similar assemblages. This study provides a repeatable framework on which future studies can assess how future changes in the EAC may influence the distribution of seabird assemblages.

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\~18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Figure** \@ref(fig:fig1-study-area)**a**; @kessler2013; @ganachaud2014). The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC flows polewards and is primarily associated with the shelf break, but at \~32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension' (@oke2019a; often termed the 'Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Figure** \@ref(fig:fig1-study-area)**a**), carrying eddy-like features as a transient current [@oke2019].

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022]; the full list of packages can be found in the Supplementary Material. Code and the modelling dataset (see \ref{sec:modellingdatasets}) are available in **(Zenodo repo)**.

@@

The spatial and temporal persistence of oceanographic features such as eddy kinetic energy ('eke') and thermal fronts (sea surface temperature gradient: 'sst_grad') may influence the movements and distributions of seabirds by offering a higher likelihood of finding food [@weimerskirch2007; @scales2014a]. Thus, 'eke' and 'sst_grad' environmental layers were calculated before extracting environmental data for each seabird record. For both 'eke' and 'sst_grad', we used their original cell spatial resolution (**Table** \@ref(tab:table1)) and calculations were done layer-by-layer to match seabird data on the finest spatial scale possible. Eddy kinetic energy was calculated as $0.5 (u'² + v'²)$, where $u' = u - \bar{u}$ and $v' = v - \bar{v}$ are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as $\bar{u}$ and $\bar{v}$ [@sun2022]. The spatial 'sst_grad' for each cell was calculated by averaging the maximum value for each neighbouring cell (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as $NS² + WE²$ [@burrows2011]. In addition, we calculated climatic seasonal average layers for 'eke' and 'sst_grad' based on the 11-year data for 2011--2021 as a proxy for persistent oceanographic features, and the 'eke' standard deviation over the same period (named with the prefix 'clim\_' in **Table** \@ref(tab:table1)). Seabed slope was calculated using 'raster::terrain()' [@raster], and distance from the coast was calculated through 'sf::st_distance()' [@sf] using Australia and Tasmania as the mainland (i.e. ignoring smaller islands). Then, for each seabird record, the most proximate value for each environmental layer (original or calculated) was extracted using built-in functions from 'raster' 3.5-21, 'terra' 1.6-7, and 'sf' 1.0-8 packages [@sf; @raster; @terra].

@@

We used the mixture model Region of Common Profiles (RCP) to identify and describe seabird species assemblages. RCP is a one-stage bioregionalization approach, as it identifies bioregions by specifying a statistical model that simultaneously relates response (multi-species data) and predictor (environmental data) [@woolley2020; @hill2020]. Given a set of environmental variables, the model can predict assemblage probabilities for sites where no biotic (observational) data exist. As a model-based approach, RCPs can be compared by information criteria, present uncertainty around the results, and check for model assumptions through assessing distribution of residuals [@warton2015; @hill2020].

The model is specified in terms of assemblages and site-specific probabilities. There are $K$ assemblages (i.e. RCPs), $K = (1, ..., k)$, that are common across the study area. The outcome variable is a vector $y_{i} = (y_{i1}, ... y_{iS})$, that gives the count or presence-absence of each of the $S$ species at site $i$. Depending on whether we model the counts or presence-absence data, the assemblages are defined in terms of the expected frequency or relative frequency of the species, respectively. An output of the model is the estimated probability that site $i$ has each of the $K$ possible assemblages. RCPs are a multivariate adaptation of a mixture-of-experts model; as such, the mixing proportions (assemblage structure) vary according to covariates (predictors) [@foster2013]. However, the final estimated assemblage structure is assumed to be constant for all sites belonging to the $K$ assemblages [@foster2013]. As such, RCPs can be directly interpreted as assemblages, and their species profiles as the assemblage structure.

@@

Due to seabird migration and dispersion, seasons could act as 'sampling artifacts' [@foster2017], thus biasing results [@ponti2023]. In light of the difference among seasons in the numbers of species that met our occurrence threshold (see \ref{sec:modellingdatasets}), we decided to use RCP models by season and data type (presence-absence and abundance). RCP models were used to relate the response variables (i.e. presence-absence or abundance of each seabird species in each grid cell) to the independent predictor covariates (i.e. averaged values of environmental variables in each cell). Presence-absence data were modelled based on a Bernoulli model, whereas abundance data were based on a Negative Binomial model. RCPs were fitted using the 'ecomix' 1.0.0 package [@ecomix], with the following workflow: (i) find the best number of groups ($K$) that describes the data, (ii) re-fit the model by fixing $K$ and selecting the optimal combination of covariates, (iii) check residuals, (iv) bootstrap the best model to estimate uncertainty around parameter estimates, and (v) produce results.

@@

A total of 10,261 seabird records was obtained during the 15 voyages, comprising 142,646 individuals from 80 species, belonging to 11 families and five orders. The five most abundant species made up 83% of total birds counted (43.0% short-tailed shearwater *Ardenna tenuirostris*, 17.2% fairy prion *Pachyptila turtur*, 14.3% wedge-tailed shearwater *Ardenna pacifica*, 5.2% sooty tern *Onychoprion fuscatus*, 3.2% shy albatross *Thalassarche cauta*). The five most frequently encountered species were *A. pacifica* (frequency of occurrence = 35.3%), *T. cauta* (33.2%), brown booby *Sula leucogaster* (25.4%), red-footed booby *S. sula* (25.4%) and black-browed albatross *T. melanophris* (25%) (**Figure S1**). An increase in species richness towards the south of the study region was clear in all but the winter season (**Figure S2**). Noticeably higher counts were made around Tasmania, possibly due to the 'colony effect' and/or the presence of Southern Ocean species (**Figure S2**). However, one cell in the Coral Sea had counts of the same order of magnitude as off Tasmania in summer.

@@

For all models, two RCPs was the optimal number of seabird groups, except for the presence-absence-autumn model, which indicated three RCPs as the best fit (**Figure S3**). The best fitting models generally had one or two covariates selected, although the abundance-autumn model had five covariates (**Table** \@ref(tab:table2)). The best models showed a good fit with no apparent deviations from the assumptions (**Figure S4**). All the best models selected 'sst' or 'sss', as essential for characterising water masses. The thresholds for defining RCP membership were 18--21°C for 'sst' and 35.1 PSU for 'sss' (**Figure** \@ref(fig:fig2-partial-plots), **Figure S5**), similar to the characteristics used to to define the EAC. Furthermore, most of the best models selected 'bat', suggesting a gradient between 'coastal' and 'oceanic' taxa based on bathymetry (**Figure S5**). Of particular note, proxies for persistent oceanographic features that may aggregate food resources (the 'climatic' variables) for seabirds were selected only in the autumn and spring abundance models.

@@

Our analyses clearly revealed a 'northern' and a 'southern' assemblage of seabirds off eastern Australia within the EAC. There is a noticeable link with the EAC dynamics: the boundaries of RCPs occurred at \~34°S, near the average separation latitude (\~32°S) of the EAC and its eastward extension into the Tasman Sea [@oke2019]. However, the 'northern/southern' assemblage pattern identified herein does not agree with previously-proposed large-scale regionalizations, such as the pelagic Longhurst Provinces [@longhurst2007], the Marine Ecoregions of the World [MEOW; @spalding2007], or the Large Marine Ecosystems [LME; @sherman1994]. The MEOW schema was based on coastal and shelf biota and was meant to represent habitats to the 200 m isobath, which clearly does not encompass the entirety of seabird habitats. The LME schema is characterised by physiographical, hydrological and productivity data but has a strong focus on practical management of areas and thus there is a political component embedded within it, which does not align with seabird distributions. @longhurst2007 designated only one biogeochemical province for eastern Australia based on productivity patterns, which may work well at larger (macro) scales, but appears to be oversimplified for the EAC.

@@

While remote tracking studies are advancing our understanding on how individual seabirds use their environment at fine scales [@tremblay2009], they fail to perceive how species assemblages interact with the environment and each other. Therefore, using at-sea counts of seabirds is a well established method to capture these relationships [@ainley2012; @woehler2003; @hyrenbach2007]. Further, the spatial scales on which assemblages relate to their environment is variable and species-specific [@hunt1987; @ribic1997; @scales2017]. Understanding the biology of the studied group is thus essential to set the appropriate scales of analyses. Our choice of 1° latitude $\times$ 1° longitude was mainly determined by data availability reflecting survey effort, and the trade-off between the number of species per grid/season (see also \ref{sec:limitations}). Nonetheless, the chosen spatial scale is similar to other seabird distribution analyses conducted elsewhere [@harris2007; @santora2013; @sojitra2022]. Therefore, for describing seabird assemblages at a macro-scale, the chosen resolution seems to have been a good fit, accords with studies elsewhere, and the resultant models described patterns of seabird species' composition consistent with our current understanding of species' ecologies and life history strategies.

@@

Although the results suggest consistent patterns, there are a number of caveats to our analyses. Seabird data were gathered during multi-disciplinary cruises, whose sampling designs were not specifically planned to account for spatial ecology of seabirds. As a result, there are spatial discontinuities in the data, which might have limited the ability of our models to detect the relationships between the seabird assemblages and the full spectrum of environmental predictors. Even though Australia is a hotspot for seabirds [@karpouzi2007; @ramírez2017], there is a paucity of information on seabird distributions in the region [@mott2018; @bernard2021]. Dedicated seabird surveys would help to advance seabird research in this region, and improve our understanding of the patterns described here. By increasing the amount of data, future studies could set the grid cells at higher resolution, and potentially reveal assemblages at finer spatial and temporal scales. In addition, ideally, we would have had more information within each grid cell by season (our sampling units). Nonetheless, rarefaction curves show that the sample coverage [see @chao2012] was high at the RCP level, although there is room for detecting more species in most of them (**Fig S8**).

>>>>

---
title: Seabird assemblages are linked to the major western boundary current off eastern Australia
authors:
  - name: Nicholas W. Daudt
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: nicholaswdaudt@gmail.com
    thanks: Corresponding author
  - name: Eric J. Woehler
    department: Australasian Seabird Group
    affiliation: BirdLife Australia
    location: Hobart, Australia
    email: eric.woehler@gmail.com
  - name: Matthew R. Schofield
    department: Department of Mathematics and Statistics
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: matthew.schofield@otago.ac.nz
  - name: Robert O. Smith
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: robert.smith@otago.ac.nz
  - name: Leandro Bugoni
    department: Institute of Biological Sciences
    affiliation: Universidade Federal do Rio Grande---FURG
    location: Rio Grande, RS, Brazil
    email: lbugoni@yahoo.com.br
  - name: William J. Rayment
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: will.rayment@otago.ac.nz
abstract: |
  **Results:** Results were based on 142,646 seabirds recorded, from 80 species, including albatrosses, petrels,  prions, shearwaters, boobies, and terns, among other taxa. All models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at ~34°S, near the latitude at which the East Australian Current (EAC) separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Main conclusions:** RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given ongoing poleward intensification that the EAC is experiencing, which is projected to continue over the next century, and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
keywords:
  - Assemblages
  - Biogeography
  - Community ecology
  - East Australian Current
  - Mixture models
  - Region of Common Profiles
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \newcommand{\blinenumbers}{\begin{linenumbers}}
- \newcommand{\elinenumbers}{\end{linenumbers}}
- \usepackage{pdflscape}
- \DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}} # Needed to understand unicode for greek Delta in Table 1
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
# As Table 1 was build enterely within this .Rmd file, I need the following hack to make sure the reference shows up in the References section:
nocite: |
  @huang2021
---

<!-- Note: I've changed the 'arxiv.sty' source file to include the running title in the PDF, and added commands in the YAML header too -->

@@

William J. Rayment <https://orcid.org/0000-0001-9558-5109>

@@

We thank all observers who assisted in data collection, the crews of RV *Investigator* and RV *Falkor*, and the Australasian Seabird Group (BirdLife Australia) for logistical support. The authors would like to thank Dr. Skipton N.C. Woolley, `{ecomix}` developer, who kindly helped NWD build preliminary models and gave advice in several stages of model construction. NWD is supported by a Doctoral scholarship from the University of Otago; LB is a research fellow from CNPq (310145/2022-8).

@@

**Data availability statement:** The gridded data used as the basis for all analyses are available in the archived repository along with the code to reproduce the results. The seabird raw data are currently being used for other projects and thus can not be publicly available. If you are interested in the seabird raw data, please contact Eric J. Woehler. The environmental data used in this study are freely available from the source (see Table 1 in the main text).

@@

**Results:** Results were based on 142,646 seabirds recorded, from 80 species, including albatrosses, petrels,  prions, shearwaters, boobies, and terns, among other taxa. All models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at \~34°S, near the latitude at which the East Australian Current (EAC) separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.

**Main conclusions:** RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given ongoing poleward intensification that the EAC is experiencing, which is projected to continue over the next century, and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.

@@

Seabird assemblages have been widely described and related to environmental characteristics [@ribic1997; @woehler2003; @hyrenbach2007]. Considering the hierarchical nature of oceanographic processes and their resulting features [@haury1978patterns], it is not surprising that the distribution of highly mobile seabirds are linked differently across temporal and spatial scales [@haury1978patterns; @hunt1987]. In the marine environment, large-scale oceanographic features (e.g. boundary currents and water masses) act similarly to terrestrial biomes (based on vegetation), and influence seabird assemblage structure---i.e. the occurrence of species and their relative numerical composition [@ribic1997; @hyrenbach2007; @commins2014]. In contrast, local abundance is related to meso- (10--100 km) and sub-mesoscale (1--10 km) processes that aggregate prey and therefore predators, such as fronts and eddies [@weimerskirch2007; @scales2014]. Such processes, although relatively ephemeral, are primarily driven by physical and atmospheric forces, and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur [@davoren2013; @hazen2013]. Sea surface temperatures are a key driver of seabird distributions, at individual- and assemblage-levels, operating over a range of spatial and temporal scales from macro (e.g. water masses) to fine (e.g. fronts) [@dunlop1988; @ribic1997; @quillfeldt2015]. Despite extensive surveys [e.g. @ballance1997] and tracking studies [e.g. @carneiro2020], many oceanic regions still lack information on the distributions and abundances of seabird species, and how they assemble in space and time.

@@

In this study, we aim to identify assemblages of seabird species within the EAC and adjacent areas, and relate these assemblages to environmental conditions, by using at-sea survey data and a state-of-the-art modelling technique. We use the Region of Common Profiles -- RCP [@foster2013], a mixture model capable of describing bioregions [@hill2020; @woolley2020]. RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics [@warton2015; @hill2020]. Eastern Australia supports a high diversity of seabirds and offers wide environmental gradients over a near 40° latitudinal range, from the tropics to the cool temperate. We hypothesised to find different seabird assemblages relating to different environmental conditions within the EAC. In addition, presence-absence and abundance data were used to test if the different numerical resolutions would result in the identification of similar assemblages. This study provides a repeatable framework on which future studies can assess how future changes in the EAC may influence the distribution of seabird assemblages.

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\~18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Figure** \@ref(fig:fig1-study-area)**a**; @kessler2013; @ganachaud2014). The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC flows polewards and is primarily associated with the shelf break, but at \~32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension', often termed the 'Tasman Front' [@oke2019a]. Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Figure** \@ref(fig:fig1-study-area)**a**), carrying eddy-like features as a transient current [@oke2019].

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022]; the full list of packages can be found in the Supplementary Material. Code and the modelling dataset (see Section \ref{sec:modellingdatasets}) are available in **(Zenodo)**.

@@

The spatial and temporal persistence of oceanographic features such as eddy kinetic energy ('eke') and thermal fronts (sea surface temperature gradient: 'sst_grad') may influence the movements and distributions of seabirds by offering a higher likelihood of finding food [@weimerskirch2007; @scales2014a]. Thus, 'eke' and 'sst_grad' environmental layers were calculated before extracting environmental data for each seabird record. For both 'eke' and 'sst_grad', we used their original cell spatial resolution (**Table** \@ref(tab:table1)) and calculations were done layer-by-layer to match seabird data on the finest spatial scale possible. Eddy kinetic energy was calculated as $0.5 (u'^{2} + v'^{2})$, where $u' = u - \bar{u}$ and $v' = v - \bar{v}$ are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as $\bar{u}$ and $\bar{v}$ [@sun2022]. The spatial 'sst_grad' for each cell was calculated by averaging the maximum value for each neighbouring cell (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as $NS^{2} + WE^{2}$ [@burrows2011]. In addition, we calculated climatic seasonal average layers for 'eke' and 'sst_grad' based on the 11-year data for 2011--2021 as a proxy for persistent oceanographic features, and the 'eke' standard deviation over the same period (named with the prefix 'clim\_' in **Table** \@ref(tab:table1)). Seabed slope was calculated using 'raster::terrain()' [@raster], and distance from the coast was calculated through 'sf::st_distance()' [@sf] using Australia and Tasmania as the mainland (i.e. ignoring smaller islands). Then, for each seabird record, the most proximate value for each environmental layer (original or calculated) was extracted using built-in functions from 'raster' 3.5-21, 'terra' 1.6-7, and 'sf' 1.0-8 packages [@sf; @raster; @terra].

@@

We used the mixture model Region of Common Profiles (RCP) to identify and describe seabird species assemblages. RCP is a one-stage bioregionalization approach, as it identifies bioregions by specifying a statistical model that simultaneously relates response (multi-species data) and predictor (environmental data) [@woolley2020; @hill2020]. Given a set of environmental variables, the model can predict assemblage probabilities for sites where no biotic (observational) data exist. As a model-based approach, RCPs can be compared by information criteria, present uncertainty around the results, and check for model assumptions through assessing residuals [@warton2015; @hill2020].

The model is specified in terms of assemblages and site-specific probabilities. There are $K$ assemblages (i.e. RCPs), $K = (1, ..., k)$, that are common across the study area. The outcome variable is a vector $y_{i} = (y_{i1}, ... y_{iS})$, that gives the count or presence-absence of each of the $S$ species at site $i$. Depending on whether we model the counts or presence-absence data, the assemblages are defined in terms of the expected frequency or relative frequency of the species, respectively. An output of the model is the estimated probability that site $i$ has each of the $K$ possible assemblages. RCPs are a multivariate adaptation of a mixture-of-experts model; as such, the mixing proportions (assemblage structure) vary according to covariates (predictors) [@foster2013]. RCPs can be directly interpreted as assemblages; the assemblage structure is given by the species profiles.

@@

Due to seabird migration and dispersion, seasons could act as 'sampling artifacts' [@foster2017], thus biasing results [@ponti2023]. In light of the difference among seasons in the numbers of species that met our occurrence threshold (see Section \ref{sec:modellingdatasets}), we decided to use RCP models by season and data type (presence-absence and abundance). RCP models were used to relate the response variables (i.e. presence-absence or abundance of each seabird species in each grid cell) to the independent predictor covariates (i.e. averaged values of environmental variables in each cell). Presence-absence data were modelled based on a Bernoulli model, whereas abundance data were based on a negative binomial model. RCPs were fitted using the 'ecomix' 1.0.0 package [@ecomix], with the following workflow: (i) find the best number of groups ($K$) that describes the data, (ii) re-fit the model by fixing $K$ and selecting the optimal combination of covariates, (iii) check residuals, (iv) bootstrap the best model to estimate uncertainty of parameter estimates, and (v) produce results.

@@

A total of 10,261 seabird records was obtained during the 15 voyages, comprising 142,646 individuals from 80 species, belonging to 11 families and five orders. The five most abundant species made up 83% of total birds counted (43.0% short-tailed shearwater *Ardenna tenuirostris*, 17.2% fairy prion *Pachyptila turtur*, 14.3% wedge-tailed shearwater *Ardenna pacifica*, 5.2% sooty tern *Onychoprion fuscatus*, 3.2% shy albatross *Thalassarche cauta*). The five most frequently encountered species were *A. pacifica* (frequency of occurrence = 35.3%), *T. cauta* (33.2%), brown booby *Sula leucogaster* (25.4%), red-footed booby *S. sula* (25.4%) and black-browed albatross *T. melanophris* (25%) (**Figure S1**). An increase in species richness towards the south of the study region was clear in all but the winter season (**Figure S2**). Noticeably higher counts were made around Tasmania, possibly due to the 'colony effect' and/or the presence of Southern Ocean species (**Figure S2**). However, in summer, one cell in the Coral Sea had counts of the same order of magnitude as off Tasmania.

@@

For all models, two RCPs was the optimal number of seabird groups, except for the presence-absence-autumn model, which indicated three RCPs as the best fit (**Figure S3**). The best fitting models generally had one or two covariates selected, although the abundance-autumn model had five covariates (**Table** \@ref(tab:table2)). The best models showed a good fit with no apparent deviations from the assumptions (**Figure S4**). All the best models selected sea surface temperature ('sst') or salinity ('sss'), as essential for characterising water masses. The thresholds for defining RCP membership were 18--21°C for 'sst' and 35.1 PSU for 'sss' (**Figure** \@ref(fig:fig2-partial-plots), **Figure S5**), similar to the characteristics used to to define the EAC. Furthermore, most of the best models selected bathymetry ('bat'), suggesting a gradient between 'coastal' and 'oceanic' taxa (**Figure S5**). Of particular note, proxies for persistent oceanographic features that may aggregate food resources (the 'climatic' variables) for seabirds were selected only in the autumn and spring abundance models.

@@

Our analyses clearly revealed a 'northern' and a 'southern' assemblage of seabirds off eastern Australia within the EAC. There is a noticeable link with the EAC dynamics: the boundaries of RCPs occurred at \~34°S, near the average separation latitude (\~32°S; 25--38°S) of the EAC and its eastward extension into the Tasman Sea [@oke2019]. However, the 'northern/southern' assemblage pattern identified herein does not agree with previously-proposed large-scale regionalizations, such as the pelagic Longhurst Provinces [@longhurst2007], the Marine Ecoregions of the World [MEOW; @spalding2007], or the Large Marine Ecosystems [LME; @sherman1994]. The MEOW schema was based on coastal and shelf biota and was meant to represent habitats to the 200 m isobath, which clearly does not encompass the entirety of seabird habitats. The LME schema is characterised by physiographical, hydrological and productivity data but has a strong focus on practical management of areas and thus there is a political component embedded within it, which does not align with seabird distributions. @longhurst2007 designated only one biogeochemical province for eastern Australia based on productivity patterns, which may work well at larger (macro) scales, but appears to be oversimplified for the EAC.

@@

While remote tracking studies are advancing our understanding on how individual seabirds use their environment at fine scales [@tremblay2009], they fail to perceive how species assemblages interact with the environment and each other. Therefore, using at-sea counts of seabirds is a well established method to capture these relationships [@ainley2012; @woehler2003; @hyrenbach2007]. Further, the spatial scales on which assemblages relate to their environment is variable and species-specific [@hunt1987; @ribic1997; @scales2017]. Understanding the biology of the studied group is thus essential to set the appropriate scales of analyses. Our choice of 1° latitude $\times$ 1° longitude was mainly determined by data availability reflecting survey effort, and the trade-off between the number of species per grid/season (see also Section \ref{sec:limitations}). Nonetheless, the chosen spatial scale is similar to other seabird distribution analyses conducted elsewhere [@harris2007; @santora2013; @sojitra2022]. Therefore, for describing seabird assemblages at a macro-scale, the chosen resolution seems to have been a good fit, accords with studies elsewhere, and the resultant models described patterns of seabird species' composition consistent with our current understanding of species' ecologies and life history strategies.

@@

Although the results suggest consistent patterns, there are a number of caveats to our analyses. Seabird data were gathered during multi-disciplinary cruises, whose sampling designs were not specifically planned to account for spatial ecology of seabirds. As a result, there are spatial discontinuities in the data, which might have limited the ability of our models to detect the relationships between the seabird assemblages and the full spectrum of environmental predictors. Even though Australia is a hotspot for seabirds [@karpouzi2007; @ramírez2017], there is a paucity of information on seabird distributions in the region [@mott2018; @bernard2021]. Dedicated seabird surveys would help to advance seabird research in this region, and improve our understanding of the patterns described here. By increasing the amount of data, future studies could set the grid cells at higher resolution, and potentially reveal assemblages at finer spatial and temporal scales. In addition, ideally, we would have had more information within each grid cell by season (our sampling units). Nonetheless, rarefaction curves show that the sample coverage [see @chao2012] was high at the RCP level, although there is room for detecting more species in most of them (**Figure S8**).
